#### Geodata Analysis
Moving forward, we will check out the popular routes and the details of them. 

Let's check out the data availability as there are missing data in the station input
```{r check for completion of the data}
skim_without_charts(year_trip_3$start_station_name)
skim_without_charts(year_trip_3$end_station_name)
```

```{r Drop rows with empty station information}
year_trip_drna <- subset(year_trip_3, !(is.na(year_trip_3$start_station_name) | is.na(year_trip_3$end_station_name)))
# drop rows with multiple condition

skim_without_charts(year_trip_drna)

```

Now the data of station is cleaned and there is no missing data, we will start with the most popular starting station and ending station. 
```{r calculate the popular start and end stations}
# popular start stations
start_stat_popular <- year_trip_drna %>%
  group_by(start_station_name) %>%
  summarize(No_of_rides = n()) %>%
  arrange(desc(No_of_rides))

head(start_stat_popular)

# popular end stations
end_stat_popular <- year_trip_drna %>%
  group_by(end_station_name) %>%
  summarize(No_of_rides = n()) %>%
  arrange(desc(No_of_rides))

head(end_stat_popular)
```

We will also see which route is the most popular one. 

```{r Calculate the popular route}
year_trip_drna <- year_trip_drna %>%
  unite("Route", start_station_name, end_station_name, sep = " to ", remove = FALSE) # unite can be directly use to insert a column if we were to concatenate two columns. remove=FALSE is needed if we do not wish to remove the input columns.

head(year_trip_drna)

# count the routes
year_trip_drna %>%
  count(Route, sort = TRUE)

```

Visualize the top ten route
```{r Top 10 route}
Top10_Route <- year_trip_drna %>%
  group_by(Route) %>%
  summarize(No_of_rides = n()) %>%
  arrange(desc(No_of_rides))

Top10_Route

Top10_Route[1:10,] %>% 
  ggplot() + geom_col(mapping = aes(x = Route, y = No_of_rides, position = "dodge"), fill = "darkcyan", color = "black") +
  coord_flip() +
  labs(title = "Top 10 Route",
       x = "Route",
       y = "Popularity in terms of No. of Rides") +
  geom_text(aes(x = Route, y = No_of_rides, label = No_of_rides), hjust = 1) +
  theme_bw()
# flip the graph to horizontal layout.
# why add the comma for Top10_Route[1:10,]? because it has two columns and the reason you need to add a comma is because R uses the following syntax for subsetting data frames: data[rows you want, columns you want] If you only type Top10_Route[1:10], then you’re telling R to return the first 10 rows, but you’re not telling R which columns to return. By using Top10_Route[1:10,], you’re telling R to return the first 10 rows and all of the columns in the data frame. 
```

Mapview of the top 100 popular starting stations
```{r Top 100 Start station Map}
Top_start_stat <- year_trip_drna %>%
  group_by(start_station_name, start_lat, start_lng, user_type) %>%
  summarize(no_of_rides = n()) %>%
  arrange(-no_of_rides)

Start_station <- Top_start_stat[1:100,] %>%
  mapview(
    xcol = "start_lng", 
        ycol = "start_lat",
        cex = "no_of_rides",
        alpha = 0.9, 
        crs = 4269,
        color = "green",
        grid = F, 
        legend = T,
        layer.name = "100 Most Popular Start Stations"
  )

Start_station
  
```

Mapview of the top 100 popular destination stations
```{r Top 100 Destination station map}
Top_end_stat <- year_trip_drna %>%
  group_by(end_station_name, end_lat, end_lng, user_type) %>%
  summarize(no_of_rides = n()) %>%
  arrange(-no_of_rides)

Destination_station <- Top_end_stat[1:100,] %>%
  mapview(
    xcol = "end_lng", 
        ycol = "end_lat",
        cex = "no_of_rides",
        alpha = 0.9, 
        crs = 4269,
        color = "#8b0000",
        grid = F, 
        legend = T,
        layer.name = "100 Most Popular Destination Stations"
  )

Destination_station
```

```{r viewing maps side by side}
sync (Start_station, Destination_station)
# reference is made from https://www.infoworld.com/article/3644848/astonishingly-easy-mapping-in-r-with-mapview.html
```

```{r Showing the maps side by side}
library(leaflet.extras2)
Start_station | Destination_station
```


  
  
